```{r, eval=FALSE}
url <- "https://scienceverse.org/talks/forrt-ai-conference/#/title-slide"
subtitle <- sprintf("[%s](%s)", url, url)
# make QR code
qrcode::qr_code(url) |> qrcode::generate_svg("images/qrcode.svg")
```

---
title: "Metacheck: automated, validated checks for open science practices"
subtitle: Cristian Mesquida
author: "![](images/qrcode.svg){width=300}"
format: 
  revealjs:
    logo: images/metacheck.png
    theme: [moon, style.scss]
    transition: none
    transition-speed: fast
    syntax-highlighting: zenburn
---

```{r, include = FALSE}
library(tidyverse)
library(flextable)
library(metacheck)
```

```{r defaults}
set_flextable_defaults(
  font.size = 20,
  padding.top = ,
  font.color = "white",
  border.color = "grey"
)
theme_set(theme_minimal(base_size = 17))
```

## Power

::: {style="font-size: 90%;"}
- **false positive rate**: the probability a <i class="st">test</i> concludes there is an <i class="es">effect</i> when there is no effect; Type I Error Rate
- <i class="a">alpha</i> : the false positive rate we accept for a <i class="st">test</i>
- **false negative rate**: the probability a <i class="st">test</i> concludes there is no <i class="es">effect</i> when there is one; Type II Error Rate
- **beta**: the false negative rate we accept for a test
- **true positive rate**: the probability a <i class="st">test</i> concludes there is an <i class="es">effect</i> when there is one
- <i class="power">power</i>: the true positive rate (1-beta), given a <i class="st">test</i>, <i class="es">effect</i>, <i class="ss">sample size</i>, and <i class="a">alpha</i>
:::

## Power analyses

Studies designed with high <i class="power">power</i> are more informative:

- Higher probability of detecting the effect size of interest when it exists

- A non-significant result is more informative because it is less likely to be due to insufficient power. For example, a non-significant result from a study desgined with 90% <i class="power">power</i> is more informative than one from a study designed with 50% <i class="power">power</i>

## Types of Power Analysis

:::::::: columns
::::::: {.column width="80%"}
- <i class="pt">**A Priori**</i>: calculates the <i class="ss">sample size</i> required to achieved a pre-specified <i class="power">power</i> for a pre-specified <i class="es">effect size</i>

- <i class="pt">**Sensitivity (Effect Size)**</i>: calculates the smallest <i class="es">effect size</i> required to detect a pre-specified <i class="ss">sample size</i> with a pre-specified <i class="power">power</i>

- <i class="pt">**Sensitivity (Power)**</i>: calculates <i class="power">power</i> given a pre-specified <i class="es">effect size</i> and <i class="ss">sample size</i>

- <i class="pt">**Post Hoc**</i>: calculates the <i class="power">power</i> associated with the observed <i class="es">effect size</i> and the <i class="ss">sample size</i> actually obtained
:::::::
::::::::

## Reporting

- <i class="pt">power type</i>: a priori, power/sample size sensitivity, post hoc
- <i class="power">power</i>: value between alpha and 100%
- <i class="ss">sample size</i>: number of subjects/items
- <i class="es">effect size</i>: numeric value
- <i class="esm">effect size metric</i>: e.g., Cohen's $d$, $r$, $\eta_{p}^{2}$, raw
- <i class="st">statistical test</i>: full specification
- <i class="a">alpha</i>: 0-1, often assumed to be 0.05
- <i class="sw">software</i>: e.g., G\*Power, Superpower, simulation!

## 

How would you write up this power analysis?

```{r}
#| echo: true

pwr::pwr.t.test(          
  power = 0.80,           
  d = 0.5,                
  sig.level = 0.01,       
  type = "two.sample",    
  alternative = "greater" 
)
```

## Model Reporting Example

> We conducted an <i class="pt">a priori analysis</i> to determine the sample size with <i class="power">80%</i> power to detect the smallest effect size of interest (SESOI), which was a <i class="esm">Cohen's d</i> of <i class="es">0.5</i> determined by a cost-benefit analysis of the intervention. We used the <i class="sw">pwr package (Champely, 2020)</i> for an <i class="st">independent samples t-test, assuming equal variance and using a one-tailed prediction</i>, with a critical alpha of <i class="a">0.01</i>. This determined that we would need <i class="ss">82</i> subjects in each group.

## The Reality

> A sample size of <i class="ss">22</i> participants was predetermined to yield a power of <i class="power">80%</i> and to detect an effect of comparable size to previous studies using similar independent and dependent measures (CITE).

## The Reality

> A <i class="sw">G\*Power (Version 3.1; Faul et al., 2009)</i> analysis using pilot data from 20 participants for a <i class="a">Bonferroni-corrected (for two comparisons)</i> <i class="st">one-sample t test</i> (<i class="esm">d</i> = <i class="es">0.45</i>, α = .05, p = .95) determined that we should recruit <i class="ss">67</i> participants.

## The Reality

> We decided to obtain a sample size that would allow us to detect a within-subjects difference of at least a medium effect size (<i class="esm">Cohen's d</i> \> <i class="es">0.5</i>) with <i class="power">80%</i> power, using a <i class="st">two-tailed t-test</i>. A power analysis using <i class="sw">G\*Power (Version 3.1.9.3; Faul et al., 2007)</i> indicated that this required a sample size of <i class="ss">34</i>.

## We Could Do Better {.smaller}

Only 11% of 1295 power analyses from 1937 *Psychological Science* papers published between 2014 and 2026 have complete reporting.

| Component                             | Percent Missing[^index-1] |
|:--------------------------------------|--------------------------:|
| <i class="pt">power type</i>          |                        0% |
| <i class="power">power</i>            |                       15% |
| <i class="ss">sample size</i>         |                       19% |
| <i class="es">effect size</i>         |                       32% |
| <i class="esm">effect size metric</i> |                       28% |
| <i class="st">statistical test</i>    |                       22% |
| <i class="a">alpha</i>                |                       67% |
| <i class="sw">software</i>            |                       71% |

## The solutions

::: {.fragment .fade-in-then-out}
- Increase awareness through better training
- Develop reporting guidelines
:::

::: {.fragment .fade-up}
- Automation
:::

## Metacheck

- It performs automated checks on **manuscripts** and **repositories**

::: {.fragment .fade-in-then-out}
- Manuscripts: `stat_effect_size`, `stat_nonsig`, `marginal`, `power`, etc.
:::

::: {.fragment .fade-up}
- Repositories: `repo_check`, `code_check`, `data_check`, etc.
:::

## Checking for power analysis

The `power` module

- **Regex-based approach**: only detects the type of power analysis based on regex

- **LLM-based approach**: identifies components of the power analysis

## LLM-based approach

```{r}
#| echo: FALSE
#| eval: FALSE

psychsci <- papers_load("psychsci") # Load a set of articles
llm_use(TRUE) # Set to true to enable LLM
llm_model("ollama") # LLM model
res <- module_run(psychsci[100], "power") # run power module
saveRDS(res, "res.RDS") # store for efficiency
```

```{r}
#| echo: FALSE
#| eval: TRUE
res <- readRDS("res.RDS")
```

```{r}
#| echo: TRUE
#| eval: FALSE

llm_use(TRUE)
llm_model("ollama")
res <- module_run(psychsci[110], "power")
```

```{r}
#| echo: TRUE
res$summary_table |> 
  flextable()
```

## First match

```{r}
#| echo: TRUE
res$table[1, c(
  "power_type", 
  "statistical_test", 
  "sample_size", 
  "alpha_level", 
  "power", 
  "effect_size", 
  "effect_size_metric", 
  "software"
  )] |> 
  flextable()
```

## Second match

```{r}
#| echo: TRUE
res$table[2, c(
  "power_type", 
  "statistical_test", 
  "sample_size", 
  "alpha_level", 
  "power", 
  "effect_size", 
  "effect_size_metric", 
  "software"
  )] |> 
  flextable()
```

## The schema

```json
{
  "$schema": "https://json-schema.org/draft/2020-12/schema",
  "$id": "https://scienceverse.org/schema/power.json",
  "title": "Power Analyses",
  "description": "A power analysis.",
  "type": "object",
  "properties": {
    "text": {
      "description": "The specific text that contains all of the information used to determine this object's properties.",
      "type": ["string", "null"]
    },
    
    "power_type": {
      "description": " ... ",
      "type": ["string", "null"],
      "enum": ["apriori", "sensitivity", "posthoc", null]
    },

    "statistical_test": {
      "description": "The statistical test used. Use null if unclear.",
      "type": ["string", "null"],
      "enum": [
        "paired t-test",
        "unpaired t-test",
        "one-sample t-test",
        "1-way ANOVA",
        "2-way ANOVA",
        ...
        "other",
        null
      ]
... 
    }
```

## Validation
- Conducted on 250 articles from Psychological Science
- Created an over-inclusive ground-truth dataset
- Manually coded whether each matched statement was a power analysis (0/1)
- Manually coded:
    - <i class="pt">power type</i>
    - <i class="power">power</i>
    - <i class="ss">sample size</i>
    - <i class="es">effect size</i>
    - <i class="esm">effect size metric</i>
    - <i class="st">statistical test</i>
    - <i class="a">alpha</i>
    - <i class="sw">software</i>

## Validation metrics

| Component                             | Agree (n) | Accuracy[^index-1] |
|---------------------------------------|----------:|-------------------:|
| <i class="pt">power type</i>          | 165       | 80%      |
| <i class="st">statistical test</i>    | 159       | 77%      |
| <i class="ss">sample size</i>         | 173       | 84%      |
| <i class="a">alpha</i>                | 201       | 97%      |
| <i class="power">power</i>            | 194       | 94%      |
| <i class="esm">effect size metric</i> | 193       | 93%      |
| <i class="es">effect size</i>         | 189       | 91%      |
| <i class="sw">software</i>            | 192       | 93%      |

: Percentage of matched components for which the observed and ground-truth values agree.

**Note.** Of the 224 observed extraction rows and 225 ground-truth rows, 207 were successfully matched. Accuracy was calculated as the number of agreements for each component divided by the number of matched cases (207), multiplied by 100.



[^index-1]: Preliminary analysis using [Metacheck](https://scienceverse.github.io/metacheck), from Mesquida, Lakens & DeBruine (in prep)

